Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-22 paper.tex
Add OFH to intro
Closes #153

\section{Background}
Variant Call Format (VCF) is the standard format for interchanging genetic
variation data, encoding information about 
DNA sequence polymorphisms among a set of samples with associated 
quality control metrics and metadata~\citep{danecek2011variant}. 
Originally defined specifically as a text file,
it has been refined and standardised~\citep{rehm2021ga4gh} and the 
underlying data-model is now deeply embedded in bioinformatics practice.
Dataset sizes have grown explosively since the introduction of 
VCF as part of 1000 Genomes project~\citep{10002015global},
with Biobank scale initiatives such as 
Genomics England~\cite{turnbull2018100},
UK Biobank~\citep{bycroft2018genome,backman2021exome,halldorsson2022sequences,uk2023whole},
and the All of Us research program~\citep{all2024genomic}
collecting genome sequence data for hundreds of thousands of humans.
Large genetic variation datasets are also being generated for other organisms
and a variety of purposes including 
agriculture~\citep{ros2020accuracy,wang2023rice}, 
conservation~\citep{shaffer2022landscape}
and infectious disease surveillance~\citep{hamid2023pf7}.
VCF's simple text-based design and widespread
support~\cite{garrison2022spectrum} makes it an 
excellent archival format, but it is an inefficient basis for analysis.
Methods that require efficient access to genotype data
either require conversion to the
PLINK~\cite{purcell2007plink,chang2015second} 
or BGEN~\citep{band2018bgen} 
formats~\citep[e.g.][]{yang2011gcta,mbatchou2021computationally,loh2015efficient}
or use bespoke binary formats 
that support the required access patterns~\citep[e.g.][]{
% Uses custom "bref3" format,
% https://faculty.washington.edu/browning/beagle/bref3.24May18.pdf
browning2018one, 
% .samples Zarr format
kelleher2019inferring,
% Has a "xcftools" package, but it still looks pretty experimental
hofmeister2023accurate}.
While PLINK and BGEN formats are more efficient to access than VCF, neither
can accommodate the full flexibility of the VCF data model and conversion
is lossy.
PLINK's approach of storing the genotype matrix in uncompressed
packed-binary format provides efficient access to genotype 
data, but file sizes are substantially larger than 
the equivalent compressed VCF (see Fig~\ref{fig-data-storage}).
For example, at two bits 
per diploid genotype, the full genotype matrix for the GraphTyper SNP dataset 
in the 500K UKB WGS data~\citep{uk2023whole} is 116 TiB.
% 1,037,556,156 SNPs x  490,640 samples
% humanize.naturalsize(1_037_556_156 * 490_640 / 4, binary=True)
% '115.7 TiB'

>>>>

\section{Background}
Variant Call Format (VCF) is the standard format for interchanging genetic
variation data, encoding information about 
DNA sequence polymorphisms among a set of samples with associated 
quality control metrics and metadata~\citep{danecek2011variant}. 
Originally defined specifically as a text file,
it has been refined and standardised~\citep{rehm2021ga4gh} and the 
underlying data-model is now deeply embedded in bioinformatics practice.
Dataset sizes have grown explosively since the introduction of 
VCF as part of 1000 Genomes project~\citep{10002015global},
with Biobank scale initiatives such as 
Genomics England~\cite{turnbull2018100},
UK Biobank~\citep{bycroft2018genome,backman2021exome,halldorsson2022sequences,uk2023whole},
Our Future Health~\citep{ofhmain_2025},
and the All of Us research program~\citep{all2024genomic}
collecting genome sequence data for hundreds of thousands of humans.
Large genetic variation datasets are also being generated for other organisms
and a variety of purposes including 
agriculture~\citep{ros2020accuracy,wang2023rice}, 
conservation~\citep{shaffer2022landscape}
and infectious disease surveillance~\citep{hamid2023pf7}.
VCF's simple text-based design and widespread
support~\cite{garrison2022spectrum} makes it an 
excellent archival format, but it is an inefficient basis for analysis.
Methods that require efficient access to genotype data
either require conversion to the
PLINK~\cite{purcell2007plink,chang2015second} 
or BGEN~\citep{band2018bgen} 
formats~\citep[e.g.][]{yang2011gcta,mbatchou2021computationally,loh2015efficient}
or use bespoke binary formats 
that support the required access patterns~\citep[e.g.][]{
% Uses custom "bref3" format,
% https://faculty.washington.edu/browning/beagle/bref3.24May18.pdf
browning2018one, 
% .samples Zarr format
kelleher2019inferring,
% Has a "xcftools" package, but it still looks pretty experimental
hofmeister2023accurate}.
While PLINK and BGEN formats are more efficient to access than VCF, neither
can accommodate the full flexibility of the VCF data model and conversion
is lossy.
PLINK's approach of storing the genotype matrix in uncompressed
packed-binary format provides efficient access to genotype 
data, but file sizes are substantially larger than 
the equivalent compressed VCF (see Fig~\ref{fig-data-storage}).
For example, at two bits 
per diploid genotype, the full genotype matrix for the GraphTyper SNP dataset 
in the 500K UKB WGS data~\citep{uk2023whole} is 116 TiB.
% 1,037,556,156 SNPs x  490,640 samples
% humanize.naturalsize(1_037_556_156 * 490_640 / 4, binary=True)
% '115.7 TiB'
